Quantum Interval-Valued Probability: Contextuality and the Born Rule

### 2017-10-30 prop_letter.tex
prop_letter.tex \section{The Born rule and Gleason's Theorem} Updating..

When combining non-disjoint events, there is a further subtlety whose
resolution will give us the final general condition for IVPMs. For
general, not necessarily disjoint, events $E_1$ and $E_2$, we have:
\begin{equation}
\mu(E_1\cup E_2) + \mu(E_1\cap E_2) \subseteq \mu(E_1) + \mu(E_2)
\label{eq:classicalconvex}
\end{equation}
which is a generalization of the classical inclusion-exclusion
principle that uses $\subseteq$ instead of $=$ for the same reason as
before. The new condition, known as
\emph{convexity}~\cite{Shapley1971,GilboaSchmeidler1994,NgMoYeh1997,Marinacci1999,Marinacci_2004,Grabisch2016},
reduces to the previously
motivated condition when the events are disjoint, i.e., when
$\mu(E_1\cap E_2) = 0$.

@@

This theorem follows directly from our
correspondence~\cite{HOSTunpublished} and Shapley's
theorem~\cite{Shapley1971,GilboaSchmeidler1994,NgMoYeh1997,Marinacci_2004,Grabisch2016},
which proves that a classical IVPM always contains at least one state $\rho$
that is consistent with every event in $\events_{\mathrm{C}}$. In contrast, as we
illustrate in Sec.~\ref{sec:Gleason}, the set of states
$\coreBorn\left(\bar{\mu},\events'\right)$ that is consistent with
some arbitrary (not necessarily commuting) events $\events'$ and a
particular QIVPM $\bar{\mu}$ may be \emph{empty}.

@@

\noindent Intuitively the expectation value of an observable relative
to a QIVPM $\bar{\mu}$ lies between two possible outcomes, which
themselves lie between the minimum and maximum bounds of the
probability intervals associated with each state~$\rho$ that is
consistent with~$\bar{\mu}$ on every projector in the spectral
decomposition of the observable. If $\bar{\mu}$ is a conventional
(Born) probability measure induced by a state $\rho$, then the Born
rule probability induced by every state in
$\coreBorn\left(\bar{\mu},\events'\right)$ will be
$\mu_{\rho}^{\mathrm{B}}$ and the interval collapses to a point, thus
reducing the definition to that of
Eq.~(\ref{eq:quantum-expectation})~\citep{HOSTunpublished}.
% For example, we can extend one of the fragments of valid probability
% measures~$\mu_{1}$ defined in Table~\ref{tab:quantum-probability-measure}
% to a QIVPM~$\bar{\mu}_{3}$:
% \begin{equation}
% \begin{aligned}\bar{\mu}_{3}(\proj{0}) & =\left[\tfrac{1}{2},\tfrac{1}{2}\right] & \bar{\mu}_{3}(\proj{2}) & =\left[0,0\right]\\
% \bar{\mu}_{3}(\proj{1}) & =\left[\tfrac{1}{2},\tfrac{1}{2}\right] &  & \ldots\textrm{ etc.}
% \end{aligned}
% \label{eq:Born-QIVPM}
% \end{equation}
% For the same observable~$\mathbf{O}=\proj{0}+2\,\proj{1}+3\,\proj{2}$,
% the only state~$\rho$ consistent with $\bar{\mu}_{3}$ on $\proj{0}$,
% $\proj{1}$, and $\proj{2}$ must be $\proj{\ps}$, and $\mu_{1}=\mu_{\proj{\ps}}^{\mathrm{B}}$.
% Therefore, the expectation value~$\expval{\mathbf{O}}_{\bar{\mu}_{3}}=\left[\expval{\mathbf{O}}_{\mu_{1}},\expval{\mathbf{O}}_{\mu_{1}}\right]=\left[1.5,1.5\right]$
% which is an interval, but not between $0$ and $1$ because it is
% not a probability of an event. 
We also note that, when restricted to commuting projectors, the
definition above is consistent with the classical notion of the
\emph{Choquet integral}~\citep{Choquet1954,GilboaSchmeidler1994,Marinacci_2004,Grabisch2016}
which is used to calculate the expectation value of random variables
as a weighted average~\citep{HOSTunpublished}.

@@

The equivalence between Statement~\ref{enu:iota-subject-to} and
\ref{enu:iota-mu-CIVPM} is a well-known classical
result~\citep{Choquet1954,NgMoYeh1997,Marinacci_2004}.  For the
quantum case, Statement~\ref{enu:iota-subject-to} implies
\ref{enu:iota-mu-QIVPM} as we have outlined above. The rest of the
proof is to show that if $\ultramodular$ is not ultramodular, then for
some quantum probability measure $\mu$, the composite
$\ultramodular\circ\mu$ might not be a QIVPM. Suppose there are three
particular numbers~$x_{0}$, $x_{1}$, and $x_{2}\in\left[0,1\right]$
such that $x_{0}+x_{1}+x_{2}\in\left[0,1\right]$, but they don't
satisfy Eq.~(\ref{eq:iota-Inclusion}). Consider the state:
\begin{equation}
\begin{aligned}\rho={} & x_{0}\proj{0}+x_{1}\proj{1}+x_{2}\proj{2}\\
 & +\left(1-x_{0}-x_{1}-x_{2}\right)\proj{3}\,.
\end{aligned}
\end{equation}
The induced map using the Born rule and clustered by $\ultramodular$,
$\ultramodular\circ\mu_{\rho}^{\mathrm{\mathrm{B}}}$ fails to satisfy
Eq.~(\ref{eq:QuantumInterval-valuedProbability-Inclusion}) when
$P_{0}=\proj{0}+\proj{2}$ and $P_{1}=\proj{1}+\proj{2}$. In other
words, this induced map fails to be a QIVPM.

>>>>

When combining non-disjoint events, there is a further subtlety whose
resolution will give us the final general condition for IVPMs. For
general, not necessarily disjoint, events $E_1$ and $E_2$, we have:
\begin{equation}
\mu(E_1\cup E_2) + \mu(E_1\cap E_2) \subseteq \mu(E_1) + \mu(E_2)
\label{eq:classicalconvex}
\end{equation}
which is a generalization of the classical inclusion-exclusion
principle that uses $\subseteq$ instead of $=$ for the same reason as
before. The new condition, known as
\emph{convexity}~\cite{Shapley1971,GilboaSchmeidler1994,NgMoYeh1997,Marinacci1999,Grabisch2016},
reduces to the previously
motivated condition when the events are disjoint, i.e., when
$\mu(E_1\cap E_2) = 0$.

@@

This theorem follows directly from our
correspondence~\cite{HOSTunpublished} and Shapley's
theorem~\cite{Shapley1971,GilboaSchmeidler1994,NgMoYeh1997,Grabisch2016},
which proves that a classical IVPM always contains at least one state $\rho$
that is consistent with every event in $\events_{\mathrm{C}}$. In contrast, as we
illustrate in Sec.~\ref{sec:Gleason}, the set of states
$\coreBorn\left(\bar{\mu},\events'\right)$ that is consistent with
some arbitrary (not necessarily commuting) events $\events'$ and a
particular QIVPM $\bar{\mu}$ may be \emph{empty}.

@@

\noindent Intuitively the expectation value of an observable relative
to a QIVPM $\bar{\mu}$ lies between two possible outcomes, which
themselves lie between the minimum and maximum bounds of the
probability intervals associated with each state~$\rho$ that is
consistent with~$\bar{\mu}$ on every projector in the spectral
decomposition of the observable. If $\bar{\mu}$ is a conventional
(Born) probability measure induced by a state $\rho$, then the Born
rule probability induced by every state in
$\coreBorn\left(\bar{\mu},\events'\right)$ will be
$\mu_{\rho}^{\mathrm{B}}$ and the interval collapses to a point, thus
reducing the definition to that of
Eq.~(\ref{eq:quantum-expectation})~\citep{HOSTunpublished}.
% For example, we can extend one of the fragments of valid probability
% measures~$\mu_{1}$ defined in Table~\ref{tab:quantum-probability-measure}
% to a QIVPM~$\bar{\mu}_{3}$:
% \begin{equation}
% \begin{aligned}\bar{\mu}_{3}(\proj{0}) & =\left[\tfrac{1}{2},\tfrac{1}{2}\right] & \bar{\mu}_{3}(\proj{2}) & =\left[0,0\right]\\
% \bar{\mu}_{3}(\proj{1}) & =\left[\tfrac{1}{2},\tfrac{1}{2}\right] &  & \ldots\textrm{ etc.}
% \end{aligned}
% \label{eq:Born-QIVPM}
% \end{equation}
% For the same observable~$\mathbf{O}=\proj{0}+2\,\proj{1}+3\,\proj{2}$,
% the only state~$\rho$ consistent with $\bar{\mu}_{3}$ on $\proj{0}$,
% $\proj{1}$, and $\proj{2}$ must be $\proj{\ps}$, and $\mu_{1}=\mu_{\proj{\ps}}^{\mathrm{B}}$.
% Therefore, the expectation value~$\expval{\mathbf{O}}_{\bar{\mu}_{3}}=\left[\expval{\mathbf{O}}_{\mu_{1}},\expval{\mathbf{O}}_{\mu_{1}}\right]=\left[1.5,1.5\right]$
% which is an interval, but not between $0$ and $1$ because it is
% not a probability of an event. 
We also note that, when restricted to commuting projectors, the
definition above is consistent with the classical notion of the
\emph{Choquet integral}~\citep{Choquet1954,GilboaSchmeidler1994,Grabisch2016}
which is used to calculate the expectation value of random variables
as a weighted average~\citep{HOSTunpublished}.

@@

Statement~\ref{enu:iota-subject-to} implies \ref{enu:iota-mu-CIVPM}
and~\ref{enu:iota-mu-QIVPM} as we have outlined above. For the quantum case, we want
to show that if $\ultramodular$ is not ultramodular, then for
some quantum probability measure $\mu$, the composite
$\ultramodular\circ\mu$ might not be a QIVPM. Suppose there are three
particular numbers~$x_{0}$, $x_{1}$, and $x_{2}\in\left[0,1\right]$
such that $x_{0}+x_{1}+x_{2}\in\left[0,1\right]$, but they don't
satisfy Eq.~(\ref{eq:iota-Inclusion}). Consider the state:
\begin{equation}
\begin{aligned}\rho={} & x_{0}\proj{0}+x_{1}\proj{1}+x_{2}\proj{2}\\
 & +\left(1-x_{0}-x_{1}-x_{2}\right)\proj{3}\,.
\end{aligned}
\end{equation}
The induced map using the Born rule and clustered by $\ultramodular$,
$\ultramodular\circ\mu_{\rho}^{\mathrm{\mathrm{B}}}$ fails to satisfy
Eq.~(\ref{eq:QuantumInterval-valuedProbability-Inclusion}) when
$P_{0}=\proj{0}+\proj{2}$ and $P_{1}=\proj{1}+\proj{2}$. In other
words, this induced map fails to be a QIVPM. For the classical case, if
$\ultramodular$ is not ultramodular, consider the commuting subspace of
event~$\events_{\mathrm{C}}$ generated by $\proj{0}$, $\proj{1}$, $\proj{2}$,
and $\proj{3}$, and then state~$\rho$ defined previously. Then, restricting
$\mu_{\rho}^{\mathrm{\mathrm{B}}}$ to $\events_{\mathrm{C}}$,
$\mu_{\rho}^{\mathrm{\mathrm{B}}}|_{\events_{0}}$, gives a classical probability
measure $\mu:\events_{\mathrm{C}}\rightarrow\left[0,1\right]$. By the same
reason as the quantum case, the induced map $\ultramodular\circ\mu$ fails to be
a classical IVPM.
